### 2024-09-23 paper/paper.qmd
Added 1 reference pertaining to the benefits of trees in cities.
Added 3 new graphs to Quarto, and added 2 paragrahs, one in the Introduction, and one in Data

---
title: "My title"
subtitle: STA304 Paper 1 Submission
author: 
  - Veyasan Ragulan
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(mapping = aes(x = botanical_name)) +
  geom_bar() +
  theme_minimal() +
  labs(x = "Tree species",
       y = "Number of trees planted")
```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

\newpage

# References

### 2024-09-23 paper/paper.qmd
Forgot to commit these two from last commit, oops!

@@
   - Another author
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at: https://github.com/Veyasan1/STA3014-Paper1"
 date: today
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
-
-The remainder of this paper is structured as follows. @sec-data....
+Urban spaces are often void of nature, whether it be grass fields, blooming flowers, or teaming and diverse natural wildlife.
+One simple way to tackle this is to incorporate nature within the urban landscape, such as with trees.
+Trees are a vital plant in the world, they take in carbon dioxide and provide oxygen, a key component in the air we breath.
+Urban areas are known for the excess in carbon dioxide, which only makes it seem natural that trees can live and perhaps thrive in such environments.
+Trees can also alter the way wind affects pedestrians below (@beptree).
+It should be no suprise then, that many cities around the world, including Toronto, have incorporated trees into their landscape, in busy downtown centres, to sleepy suburbs, and everywhere in between.
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
-
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
-
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
-
-Talk more about it.
+The data comes from OpedData Toronto, an open source portal containng datasets carried out at the municipal level (@opendata).
+This particular dataset is called Street Tree Data, and focuses on city-owned trees located on roads.
+This information would be most likely used by city planners and road maintenance, ensuring snow plows and street dusters don't impede on any trees that line the roads.
+It is important to not that there may be some privately owned trees listed among the municipal trees, as they may be of interest with regards to road maintenance or other city services.
 
@@
 ```{r}
-#| label: fig-planes
+#| label: fig-topspecies
 #| fig-cap: Relationship between wing length and width
@@
 
-analysis_data |> 
-  ggplot(mapping = aes(x = botanical_name)) +
-  geom_bar() +
+top_species_count <- analysis_data %>%
+  count(botanical_name, sort = TRUE) %>%
+  head(10)
+
+top_species_count |> 
+  ggplot(aes(y = botanical_name, x = n)) +
+  geom_col() +
   theme_minimal() +
-  labs(x = "Tree species",
+  labs(x = "Planted Trees",
+       y = NULL)
+```
+Talk way more about it.
+
+```{r}
+#| label: fig-ward
+#| fig-cap: Relationship between wing length and width
+#| echo: false
+#| warning: false
+#| message: false
+
+analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+
+ward_count <- analysis_data %>%
+  count(ward, sort = TRUE) %>%
+  head(10)
+
+ward_count |> 
+  ggplot(aes(x = ward, y = n)) +
+  geom_col() +
+  theme_minimal() +
+  labs(x = "Ward",
        y = "Number of trees planted")
@@
 
-Talk way more about it.
+```{r}
+#| label: fig-speciesward
+#| fig-cap: Relationship between wing length and width
+#| echo: false
+#| warning: false
+#| message: false
 
-# Model
+analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
 
-The goal of our modelling strategy is twofold. Firstly,...
+top_species_ward_count <- analysis_data %>%
+  filter(botanical_name == "Acer platanoides") %>%
+  count(ward, sort = TRUE) %>%
+  head(10)
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+top_species_ward_count |> 
+  ggplot(aes(x = ward, y = n)) +
+  geom_col() +
+  theme_minimal() +
+  labs(x = "Ward",
+       y = "Number of trees planted")
+```
+```{r}
+#| label: fig-trunk
+#| fig-cap: Relationship between wing length and width
+#| echo: false
+#| warning: false
+#| message: false
 
-## Model set-up
+analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+max_dbh <- analysis_data %>%
+  filter(dbh_trunk == 815)
 
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
+max_dbh
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
+analysis_data |> 
+  ggplot(mapping = aes(x = dbh_trunk)) +
+  geom_histogram() +
+  theme_minimal() +
+  labs(x = "DBH Diameter",
+       y = "Tree Count")
+```
 

### 2024-09-23 paper/paper.qmd
Added onto data in the papaer, as well as the introduction, added replications of paper's graphs in relications

@@
 
+One way cities have incorportated trees into the city are roadside trees.
+Trees line many of the cities highways, arterials, collectors, and small roads.
+Large quantaties of trees lined against highways can act as a natural sound barrier, INSERT STATISTIC WITH SORUCE HERE.
+Even in cases where trees are more sporadic, they act as
+
 # Data {#sec-data}
 
-The data comes from OpedData Toronto, an open source portal containng datasets carried out at the municipal level (@opendata).
+The data comes from OpenData Toronto, an open source portal containng datasets carried out at the municipal level (@opendata).
 This particular dataset is called Street Tree Data, and focuses on city-owned trees located on roads.
 This information would be most likely used by city planners and road maintenance, ensuring snow plows and street dusters don't impede on any trees that line the roads.
-It is important to not that there may be some privately owned trees listed among the municipal trees, as they may be of interest with regards to road maintenance or other city services.
+It is important to note that there may be some privately owned trees listed among the municipal trees, as they may be of interest with regards to road maintenance or other city services.
+
+Street Tree Data contains around 32000 observations, each observation being an individual tree.
+Each observation has a general id for data analysis, as well as an structure id, telling us if the tree is part of an existing structure or building.
+Next is the location of the tree, using the nearest parcel delivery address, nearest cross-junction, and which ward it is located in.
+Street Tree contains both the botanical name and the common english name, however we have decided to use only the botanical name to avoid confusion between similar sounding trees.
+Common english names will be provided for species of key interest.
+Finally, there is a measurement called dbh_trunk.
+This is a standard method for measuring the diameter of a tree.
+The method asks record takeres to read the circumference of the tree from 1.3m off the ground it is rooted in.
+This measurement has been known to be vaguage, especially when it comes to wawrped or tipping trees, making the exact measurements made on each tree inaccurate at best.
 

### 2024-09-24 paper/paper.qmd
Added Discussion to 2 points in paper.qmd. Added tests of is na on ward and botanical name in the test script. Updated references with new sources, added figures to replication script

@@
 ---
-title: "My title"
+title: "An abundance of invasive trees in Toronto and a lack of trees in the streets of Toronto's core. "
 subtitle: STA304 Paper 1 Submission
@@
   - Veyasan Ragulan
-  - Another author
 thanks: "Code and data are available at: https://github.com/Veyasan1/STA3014-Paper1"
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Tree- lined streets are an exceptional way to add nature to urban envirnoments.
+Using the Street Tree data from OpenData Toronto, this paper examines the patterns and trends in Toronto's street trees.
+Tree species count, count of trees by ward, and distribution of tree-diameters were analysed.
+Uncovered was a abundance of a non-native tree species, and a lack of trees on the streets of Toronto's core"
 format: pdf
@@
 library(tidyverse)
-library(palmerpenguins)
 ```
@@
 
-Urban spaces are often void of nature, whether it be grass fields, blooming flowers, or teaming and diverse natural wildlife.
+Urban spaces can be void of vast expanses of nature, whether it be grass fields, blooming flowers, or teaming and diverse natural wildlife.
 One simple way to tackle this is to incorporate nature within the urban landscape, such as with trees.
@@
 Urban areas are known for the excess in carbon dioxide, which only makes it seem natural that trees can live and perhaps thrive in such environments.
-Trees can also alter the way wind affects pedestrians below (@beptree).
+One way cities have incorportated trees into the city are roadside trees.
+Trees line many of the cities highways, arterials, collectors, and small roads.
+Lines of trees lined alongside highways can act as a natural sound barrier by reflecting, scattering and absoring nearby noises (@aais).
+Tree-lined streets can also alter the way wind affects pedestrians below them, creating a more hospitable enviroment to promote active transportation such as walking or biking (@beptree).
 It should be no suprise then, that many cities around the world, including Toronto, have incorporated trees into their landscape, in busy downtown centres, to sleepy suburbs, and everywhere in between.
 
-One way cities have incorportated trees into the city are roadside trees.
-Trees line many of the cities highways, arterials, collectors, and small roads.
-Large quantaties of trees lined against highways can act as a natural sound barrier, INSERT STATISTIC WITH SORUCE HERE.
-Even in cases where trees are more sporadic, they act as
+In this paper we look at street tree data from Toronto's roadside trees, and discern any patterns or observations to suggest further investigation.
+We first look at the most populous tree species in Toronto, an the ward with the most trees.
+We will also look at tree diameter, which can indicate the approximate age of trees in the city, and look for anomalies and interesting summaries there.
+Finally we look at the least popoulous tree species and figure out where Toronto's native trees stack up.
+
+Section 2 will outline the source of this data.
+Section 3 covers the graphs depicting interesting findings and trends.
+Section 4 is where discussion will be made about those findings and trends, as well as theories to explain them.
+Finally, section 4 discusses any weakness and limitations that can be considered for another report.
 
@@
 
-The data comes from OpenData Toronto, an open source portal containng datasets carried out at the municipal level (@opendata).
+The data comes from OpenData Toronto, an open source portal containing datasets carried out at the municipal level (@opendata).
 This particular dataset is called Street Tree Data, and focuses on city-owned trees located on roads.
@@
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
+# Results
 ```{r}
-#| label: fig-topspecies
-#| fig-cap: Relationship between wing length and width
 #| echo: false
@@
 analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+```
+
+```{r}
+#| label: fig-topspecies
+#| fig-cap: Street tree count of 10 most popoulus tree species in Toronto up to September 2024
+#| echo: false
+#| warning: false
+#| message: false
 
@@
 ```
-Talk way more about it.
+
+The most populous species by far is the Acer platanoides, commonly known as the Norwegian Maple.
+It is interesting how much more prevalent it is compared to any of the other top 10 tree species, at over 3000 recorded individuals.
 
@@
 #| label: fig-ward
-#| fig-cap: Relationship between wing length and width
+#| fig-cap: Street tree count by Toronto wards up to September 2024
 #| echo: false
@@
 
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
 ward_count <- analysis_data %>%
-  count(ward, sort = TRUE) %>%
-  head(10)
+  count(ward, sort = TRUE)
 
@@
 ```
+Ward 2 takes the top spot for tree count, with Ward 3 trailing behind it.
+These wards are both in Toronto's west end, Etobicoke Centre, and Etobicoke- Lakeshore.Ward 13 and Ward 10 have the least amount of trees, being Spadina-Fort York and Toronto Centre.
 
 ```{r}
-#| label: fig-speciesward
-#| fig-cap: Relationship between wing length and width
+#| label: fig-trunk
+#| fig-cap: Distribution of Distance-at-Breast-Height diameter of all trees in Toronto up to September 2024
 #| echo: false
@@
 
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-top_species_ward_count <- analysis_data %>%
-  filter(botanical_name == "Acer platanoides") %>%
-  count(ward, sort = TRUE) %>%
-  head(10)
-
-top_species_ward_count |> 
-  ggplot(aes(x = ward, y = n)) +
-  geom_col() +
-  theme_minimal() +
-  labs(x = "Ward",
-       y = "Number of trees planted")
-```
-```{r}
-#| label: fig-trunk
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
-
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
-
-max_dbh <- analysis_data %>%
-  filter(dbh_trunk == 815)
-
-max_dbh
-
 analysis_data |> 
-  ggplot(mapping = aes(x = dbh_trunk)) +
-  geom_histogram() +
+  ggplot(aes(x = dbh_trunk)) +
+  geom_histogram(binwidth = 10) +
   theme_minimal() +
@@
 ```
+The histogram has a very long tail to the left, indicating one or two trees having upwards of 815 cm of diameter.
 
@@
 
-## First discussion point {#sec-first-point}
+## The worrying prominence of Acer platanoides
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Figure 1 shows that Acer platanoides, otherwise known as the Norway Maple, is clearly the most populous tree in the city of Toronto, beating out any of 9 other contenders by 200 individuals.
+Acer platanoides is a well known tree used in urban environments, especially after the demise of elm trees in the 1970s (@urbanforest).
+These trees are know for ‘a vigorous early growth rate, desirable form and size, the capacity to withstand many urban impacts (e.g.
+pavement, moderate levels of pollution, dusts, and dry soils) and the abilities to transplant well, grow on a wide variety of soils, and withstand ice and snow damage better than other maples’ (@norwaymaple).
+Being planted in cities, it's seeds can travel by wind across to more wooded suburbs, where the seeds can tolerate shade for extended periods of time (@urbanforest).
+However, as it's common name suggests, it is not a native species to Canada, instead brought here by European explorers and settlers.
+While not a very invasive species, taking over just 9% of Southern Ontario's forests, it does so by harming other native plants and trees (@urbanforests).
+More attention should be cast on the spread of this tree, as it has the potential to devastate Toronto's and SOuthern Ontario's forests in the years to come.
 
-## Second discussion point
+## The supposed lack of trees in downtown Toronto
+Figure 2 shows the distribution of trees across the 25 wards of Toronto.
+Wards 2 and 3 top the rankings, at over 2000 individuals.
+This isn't a suprise as those wards are located in Etobicoke Centre and Etobicoke-Lakeshore, consiting mainly of suburbs and small pocjkets of woodland.
+On the other hand, Wards 10 and 13, Spadina-Fort York, and Toronto Centre, display a shocking lack of trees, at 500 or less.
+These wards make up a majority of Toronto's downtown core, which would explain the lack of trees.
+However, this could be a call to invest in more roadside and path side greenery.
+Implementing more trees either along big avenues such as university, or small roads servicing townhouses and condos, can help liven up what is considered by many to be a dull, drab section of the city, with enless rows of concrete and glass interspersed with concrete and brick and little else.
 
-## Third discussion point
+## DBH diameters
 
@@
 
-Weaknesses and next steps should also be included.
+Street Tree data is cumulative, meaning Urban Forestry has to manually update the status of each tree, which is an arduous task for over 30000 trees.
+As a result the data may be mismatched, with some trees having only been checked months or even years prior.
+Another limitation mentioned earlier is the inclusion of some privately-owned trees.
+There isn't a clear indicator mentioned of who owns the tree in each ovbservation.
 
-\newpage
-
-\appendix
-
-# Appendix {-}
-
-# Additional data details
+Because of the scope of the data harvested, there may be some inacuraccies.
+For example, Figure 3 is skewed extremely far to the left due to some enourmous numbers.
+There isn't a clear way of double cheking whther a dbh of 815 was actually recorded and accurate, which makes any conclusions drawn from DBH potentially inaccurate.
 

### 2024-09-24 scripts/03-test_data.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
+# Purpose: Tests the dataset for obvious errors
 # Author: Veyasan Ragulan
 # Date: 20 September 2024
 # Contact: veyasan.ragulan@mail.utoronto.ca
@@ -10,6 +10,15 @@
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
 
 #### Test data ####
+data <- read_csv("data/raw_data/simulated.csv")
+
+# Test for negative numbers
+data$DBH_TRUNK |> min() <= 0
+
+# Test for NAs
+all(is.na(data$WARD))
+all(is.na(data$BOTANICAL_NAME))
+

### 2024-09-24 scripts/05-replications.R

@@ -13,7 +13,40 @@ library(tidyverse)
 # [...UPDATE THIS...]
 
 #### Load data ####
-# [...ADD CODE HERE...]
+analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
 
+### Replicate figures ###
+
+# Figure 1: Street tree count of 10 most popoulus tree species in Toronto up to September 2024
+top_species_count <- analysis_data %>%
+  count(botanical_name, sort = TRUE) %>%
+  head(10)
+
+top_species_count |> 
+  ggplot(aes(y = botanical_name, x = n)) +
+  geom_col() +
+  theme_minimal() +
+  labs(x = "Planted Trees",
+       y = NULL)
+
+# Figure 2: Street tree count by Toronto wards up to September 2024
+ward_count <- analysis_data %>%
+  count(ward, sort = TRUE) %>%
+  head(10)
+
+ward_count |> 
+  ggplot(aes(x = ward, y = n)) +
+  geom_col() +
+  theme_minimal() +
+  labs(x = "Ward",
+       y = "Number of trees planted")
+
+# Figure 3: Distribution of Distance-at-Breast-Height diameter of all trees in Toronto up to September 2024
+analysis_data |> 
+  ggplot(mapping = aes(x = dbh_trunk)) +
+  geom_histogram() +
+  theme_minimal() +
+  labs(x = "DBH Diameter",
+       y = "Tree Count")
 
 

### 2024-09-24 paper/paper.qmd
Added references, simulations added

@@
 
-The data comes from OpenData Toronto, an open source portal containing datasets carried out at the municipal level (@opendata).
+The data comes from Urban Forestry, an organization dedicated to Toronto's urban forests.
+They work towards planting more trees, and protecting existing trees from damage due to individuals, private entities, and public entities (@urbanforestry).
 This particular dataset is called Street Tree Data, and focuses on city-owned trees located on roads.
@@
 Next is the location of the tree, using the nearest parcel delivery address, nearest cross-junction, and which ward it is located in.
-Street Tree contains both the botanical name and the common english name, however we have decided to use only the botanical name to avoid confusion between similar sounding trees.
-Common english names will be provided for species of key interest.
+Street Tree contains both the botanical name and the common English name, however we have decided to use only the botanical name to avoid confusion between similar sounding trees.
+Common English names will be provided for species of key interest.
 Finally, there is a measurement called dbh_trunk.
 This is a standard method for measuring the diameter of a tree.
-The method asks record takeres to read the circumference of the tree from 1.3m off the ground it is rooted in.
-This measurement has been known to be vaguage, especially when it comes to wawrped or tipping trees, making the exact measurements made on each tree inaccurate at best.
+The method asks record takers to read the circumference of the tree from 1.3m off the ground it is rooted in.
+This measurement has been known to be vague, especially when it comes to warped or tipping trees, making the exact measurements made on each tree inaccurate at best.
+
+Using R and R Studio
 
@@
 While not a very invasive species, taking over just 9% of Southern Ontario's forests, it does so by harming other native plants and trees (@urbanforests).
-More attention should be cast on the spread of this tree, as it has the potential to devastate Toronto's and SOuthern Ontario's forests in the years to come.
+More attention should be cast on the spread of this tree, as it has the potential to devastate Toronto's and Southern Ontario's forests in the years to come.
 
@@
 Wards 2 and 3 top the rankings, at over 2000 individuals.
-This isn't a suprise as those wards are located in Etobicoke Centre and Etobicoke-Lakeshore, consiting mainly of suburbs and small pocjkets of woodland.
+This isn't a surprise as those wards are located in Etobicoke Centre and Etobicoke-Lakeshore, consisting mainly of suburbs and small pockets of woodland.
 On the other hand, Wards 10 and 13, Spadina-Fort York, and Toronto Centre, display a shocking lack of trees, at 500 or less.
@@
 However, this could be a call to invest in more roadside and path side greenery.
-Implementing more trees either along big avenues such as university, or small roads servicing townhouses and condos, can help liven up what is considered by many to be a dull, drab section of the city, with enless rows of concrete and glass interspersed with concrete and brick and little else.
+Implementing more trees either along big avenues such as university, or small roads servicing townhouses and condos, can help liven up what is considered by many to be a dull, drab section of the city, with endless rows of concrete and glass interspersed with concrete and brick and little else.
 
 ## DBH diameters
+
+The spread of DBH diameters indicates that a small number of trees have extremely high DBH values.
+A DBH value of 800 for instance, means that the measured diameter of the tree is 8 meters!
+This is most likely an error on the part of the person in charge of inputting data at the time.
+There may have been a decimal missing from the entry, which would divide the offending dbh by 10.
+We can't be certain when the data was taken and how accurate it is however, so we leave it as is for future examination.
+The largest bin is located somewhere between 10 cm and 20 cm, indicating that the median diameter of Toronto's street trees is around that value as well.
+This suggests that most street trees are actually very young, and have not had much time to develop their trunks.
+This can be explained by initiatives to plant saplings along new or existing roads, potentially replacing older trees that had been cut down to make way for urban development.
 

### 2024-09-24 scripts/00-simulate_data.R

@@ -10,10 +10,32 @@
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(janitor)
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+set.seed(304)
 
+data <- 
+  tibble(
+    #Generating simulation for 1000 individual trees
+    "ID" = 1:1000,
+    
+    # Generating random ward locations (25) for each of the trees with replacement
+    "Ward" = sample(
+      x = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25"),
+      size = 1000,
+      replace=TRUE),
+    
+    # Generating random tree species from @streettree2021 with replacement
+    "Tree Species" = sample(
+      x = c("Acer nigrum", "Acer rubrum", "Acer saccharinum", "Acer saccharum", "Aesculus glabra", "Celtis occidentalis", "Gymnocladus dioicus", "Liriodendron tulipifera", "Ostrya virginiana", "Quercus alba", "Quercus bicolor", "Quercus macrocarpa", "Quercus shumardii", "Quercus rubra", "Tilia americana", "Acer x freemanii ‘Autumn Blaze’", "Aesculus flava", "Catalpa speciosa", "Cladrastis lutea", "Gleditsia triacanthos ‘Skyline", "Tilia americana ‘Redmond’", "Ulmus americana ‘Valley Forge’", "Aesculus hippocastanum", "Aesculus carnea ‘Briotii’", "Cercidiphyllum japonicum", "Corylus colurna", "Fagus sylvatica", "Ginkgo biloba", "Platanus x acerifolia", "Tilia cordata"),
+      size = 1000,
+      replace=TRUE),
+    
+    # Generating random dbh diameters from 1 to 100 cm for all trees with replacement
+    "DBH" = sample(1:100, 1000, replace=TRUE)
+  )
 
+#### Write_csv
+write_csv(data, file = "data/raw_data/simulated.csv")
 

### 2024-09-27 paper/paper.qmd
Fixed spelling issues, added references to R and tidyverse

--- 2024-09-24 paper/paper.qmd
@@
 title: "An abundance of invasive trees in Toronto and a lack of trees in the streets of Toronto's core. "
-subtitle: STA304 Paper 1 Submission
+subtitle: ""
 author: 
@@
 Urban areas are known for the excess in carbon dioxide, which only makes it seem natural that trees can live and perhaps thrive in such environments.
-One way cities have incorportated trees into the city are roadside trees.
+One way cities have incorporated trees into the city are roadside trees.
 Trees line many of the cities highways, arterials, collectors, and small roads.
-Lines of trees lined alongside highways can act as a natural sound barrier by reflecting, scattering and absoring nearby noises (@aais).
-Tree-lined streets can also alter the way wind affects pedestrians below them, creating a more hospitable enviroment to promote active transportation such as walking or biking (@beptree).
-It should be no suprise then, that many cities around the world, including Toronto, have incorporated trees into their landscape, in busy downtown centres, to sleepy suburbs, and everywhere in between.
+Lines of trees lined alongside highways can act as a natural sound barrier by reflecting, scattering and absorbing nearby noises (@aais).
+Tree-lined streets can also alter the way wind affects pedestrians below them, creating a more hospitable environment to promote active transportation such as walking or biking (@beptree).
+It should be no surprise then, that many cities around the world, including Toronto, have incorporated trees into their landscape, in busy downtown centres, to sleepy suburbs, and everywhere in between.
 
@@
 We will also look at tree diameter, which can indicate the approximate age of trees in the city, and look for anomalies and interesting summaries there.
-Finally we look at the least popoulous tree species and figure out where Toronto's native trees stack up.
+Finally we look at the least populous tree species and figure out where Toronto's native trees stack up.
 

### 2024-09-27 scripts/05-replications.R
Updated replications script info

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Replicated graphs from... [...UPDATE THIS...]
+# Purpose: Replicated graphs from paper.qmd
 # Author: Veyasan Ragulan
 # Date: 20 September 2024
 # Contact: veyasan.ragulan@mail.utoronto.ca
@@ -10,7 +10,6 @@
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
 
 #### Load data ####
 analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

### 2024-09-27 paper/paper.qmd
Added references

@@
 # Data {#sec-data}
+
+All collection and analysis of data was done using R, a free programming language designed for data scientists (@citeR).
+Additionally, tidycerse (@tidyverse), dplyr (@dplyr), and janitor (@janitor) have been used as packages to read, clean, and present data.
+Finally OpenData Toronto is an online database containing thousands of datasets pertaining to the City of Toronto, which is where the data for this paper was initially accessed.
 

### 2024-09-27 scripts/02-data_cleaning.R
Added code to data_clening, that replaces some wonky observation values. Updated analysis_data as a result. Also added and revised checks in test_data, making sure it runs on analysis_data instead of simulated_data

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
+# Purpose: Cleans the raw Street Tree Data collected cumulatively by Urban Forestry
 # Author: Veyasan Ragulan
 # Date: 20 September 2024
 # Contact: veyasan.ragulan@mail.utoronto.ca
@@ -9,15 +9,20 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(dplyr)
 
 #### Clean data ####
 raw_data <- read_csv("data/raw_data/raw_data.csv")
 
 cleaned_data <-
   raw_data |>
-  janitor::clean_names() |>
-  select(objectid,structid,address,streetname,ward,botanical_name,dbh_trunk) |>
-  tidyr::drop_na()
+  filter(DBH_TRUNK > 0) |> # Remove non-positive diameter readings
+  mutate(BOTANICAL_NAME = replace(BOTANICAL_NAME, BOTANICAL_NAME == "alix Ã\u0097 sepulcralis", "Salix x sepulcralis")) |> # Fix the broken botanical name
+  mutate(WARD = replace(WARD, OBJECTID == 26388, "15")) |> # Give a tree with no ward allocation its correct ward allocation
+  janitor::clean_names() |> # match case on all variable_names
+  select(objectid,structid,address,streetname,ward,botanical_name,dbh_trunk) |> # Select ids, street location, ward location, species name, and trunk diameter for analysis
+  tidyr::drop_na() # Filter out any na values
 
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")
+

### 2024-09-27 scripts/03-test_data.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Tests the dataset for obvious errors
+# Purpose: Tests the cleaned dataset for any errors
 # Author: Veyasan Ragulan
 # Date: 20 September 2024
 # Contact: veyasan.ragulan@mail.utoronto.ca
@@ -13,12 +13,21 @@ library(tidyverse)
 
 
 #### Test data ####
-data <- read_csv("data/raw_data/simulated.csv")
+data <- read_csv("data/analysis_data/analysis_data.csv")
+
+species_test <- data |>
+  filter(botanical_name == "alix Ã\u0097 sepulcralis")
+species_test
+
+# Does each tree have a corresponding ward
+ward_test <- data |>
+  filter(ward == "None")
+ward_test # Make sure this is a 0*7 tibble
 
 # Test for negative numbers
-data$DBH_TRUNK |> min() <= 0
+data$dbh_trunk |> min() <= 0
 
-# Test for NAs
-all(is.na(data$WARD))
-all(is.na(data$BOTANICAL_NAME))
+# Test for NAs in cleaned_data (should be FALSE)
+all(is.na(data$ward))
+all(is.na(data$botanical_name))
 

### 2024-09-27 paper/paper.qmd
Added references to bib file and quarto document

--- 2024-09-27 paper/paper.qmd
@@
 # Data {#sec-data}
+
+All collection and analysis of data was done using R, a free programming language designed for data scientists (@citeR).
+Additionally, tidycerse (@citeTidyverse), dplyr (@dplyr), and janitor (@citeJanitor) have been used as packages to read, clean, and present data.
+Finally OpenData Toronto is an online database containing thousands of datasets pertaining to the City of Toronto, which is where the data for this paper was initially accessed (@citeOpendatatoronto).
+
+ Rohan Alexander's book, *Telling Stories with Data* was referenced for troubleshooting and general ideas for data analysis (@tellingstories)
 

### 2024-09-27 paper/paper.qmd
Added two graphs to paper.qmd. One is a tree species count of ward 13. another is a histograph of dbh trunks without any values over 200

@@
 All collection and analysis of data was done using R, a free programming language designed for data scientists (@citeR).
-Additionally, tidycerse (@citeTidyverse), dplyr (@dplyr), and janitor (@citeJanitor) have been used as packages to read, clean, and present data.
+Additionally, tidycerse (@citeTidyverse), dplyr (@citeDplyr), and janitor (@citeJanitor) have been used as packages to read, clean, and present data.
 Finally OpenData Toronto is an online database containing thousands of datasets pertaining to the City of Toronto, which is where the data for this paper was initially accessed (@citeOpendatatoronto).
@@
 ```{r}
-#| label: fig-topspecies
+#| label: fig-1
 #| fig-cap: Street tree count of 10 most popoulus tree species in Toronto up to September 2024
@@
 
-The most populous species by far is the Acer platanoides, commonly known as the Norwegian Maple.
+Looking at @fig-1, the most populous species by far is the Acer platanoides, commonly known as the Norwegian Maple.
 It is interesting how much more prevalent it is compared to any of the other top 10 tree species, at over 3000 recorded individuals.
+Also note the similarity of the names Gleditsia triacanthos and Gleditsia triacanthos f.
+inermis 'Skyline'.
+The latter is an example of a hybrid tree, which is why it's botanical name contains the botanical name of another tree.
 
 ```{r}
-#| label: fig-ward
+#| label: fig-2
 #| fig-cap: Street tree count by Toronto wards up to September 2024
@@
 ```
+@fig-2 shows the count of trees over each ward in Toronto.
+Toronto has 25 wards, each with varying land sizes and populations.
 Ward 2 takes the top spot for tree count, with Ward 3 trailing behind it.
-These wards are both in Toronto's west end, Etobicoke Centre, and Etobicoke- Lakeshore.Ward 13 and Ward 10 have the least amount of trees, being Spadina-Fort York and Toronto Centre.
+These wards are both in Toronto's west end, Etobicoke Centre, and Etobicoke- Lakeshore.
+Ward 13 and Ward 10 have the least amount of trees, being Spadina-Fort York and Toronto Centre.
 
 ```{r}
-#| label: fig-trunk
+#| label: fig-3
+#| fig-cap: 10 most populous tree species in Ward 13 of Toronto, Ontario
+#| echo: false
+#| warning: false
+#| message: false
+
+ward13_species_count <- analysis_data %>%
+  filter(ward == "13") %>%
+  count(botanical_name, sort = TRUE) %>%
+  head(10)
+
+ward13_species_count |> 
+  ggplot(aes(y = botanical_name, x = n)) +
+  geom_col() +
+  theme_minimal() +
+  labs(x = "Planted Trees",
+       y = NULL)
+```
+
+```{r}
+#| label: fig-4
 #| fig-cap: Distribution of Distance-at-Breast-Height diameter of all trees in Toronto up to September 2024
@@
   theme_minimal() +
-  labs(x = "DBH Diameter",
+  labs(x = "DBH Diameter (cm)",
        y = "Tree Count")
 ```
-The histogram has a very long tail to the left, indicating one or two trees having upwards of 815 cm of diameter.
+@fig-4 is a histogram of the frequency of certain tree-diameters across Toronto. The histogram has a very long tail to the left, indicating one or two trees having upwards of 815 cm of diameter.
+
+```{r}
+#| label: fig-5
+#| fig-cap: Distribution of Distance-at-Breast-Height diameter of trees in Toronto up to September 2024 (excluding trees with a recorded DBH of over 200)
+#| echo: false
+#| warning: false
+#| message: false
+
+filter200_data <- analysis_data %>%
+  filter(dbh_trunk <= 200)
+
+filter200_data |> 
+  ggplot(aes(x = dbh_trunk)) +
+  geom_histogram(binwidth = 10) +
+  theme_minimal() +
+  labs(x = "DBH Diameter (cm)",
+       y = "Tree Count")
+```
 
@@
 
-Figure 1 shows that Acer platanoides, otherwise known as the Norway Maple, is clearly the most populous tree in the city of Toronto, beating out any of 9 other contenders by 200 individuals.
+@fig-1 shows that Acer platanoides, otherwise known as the Norway Maple, is clearly the most populous tree in the city of Toronto, beating out any of 9 other contenders by 200 individuals.
 Acer platanoides is a well known tree used in urban environments, especially after the demise of elm trees in the 1970s (@urbanforest).
@@
 ## The supposed lack of trees in downtown Toronto
-Figure 2 shows the distribution of trees across the 25 wards of Toronto.
+@fig-2 shows the distribution of trees across the 25 wards of Toronto.
 Wards 2 and 3 top the rankings, at over 2000 individuals.
@@
 
+# Appendix
+
+## Data Cleaning
+
+While cleaning that dataset, some inconsisten values were discovered.
+A particular species of tree, Salix x sepulcralis, had it's name misspelled in the dataset.
+This may have come from using special characters that could not be saved onto the dataset, resulting in a garbled name.
+An additional step was taken to restore the correct botanical name, using the common name as reference.
+Similarly, one observation did not have a ward location allocated.
+This was remedied by reading the globval coordinates from geometry, a varaible in the raw data file that shares the latitude and longtitude of the tree individual.
+Using these coordinates, and the address number, the general location for this tree was determined, along with its ward location.
+The ward entry for this observataion was updated accordingly.
+
 # References

### 2024-09-27 paper/paper.qmd
FInla changes made to project, undergoing final checks now

@@
 Each observation has a general id for data analysis, as well as an structure id, telling us if the tree is part of an existing structure or building.
-Next is the location of the tree, using the nearest parcel delivery address, nearest cross-junction, and which ward it is located in.
-Street Tree contains both the botanical name and the common English name, however we have decided to use only the botanical name to avoid confusion between similar sounding trees.
+Location is covered in 4 ways: first the address number (address) and streetname (streetname) of nearest postal delivery address, next is the street name and the nearest 2 cross streets (crosstreet1 and crossstreet2), third being the ward number (ward), and finally geometry containain a tuple of general latitude and longitude.
+Street Tree contains both the species name (botanical_name) and the common English name common_name), however we have decided to use only the species name to avoid confusion between similar sounding trees.
 Common English names will be provided for species of key interest.
-Finally, there is a measurement called dbh_trunk.
-This is a standard method for measuring the diameter of a tree.
-The method asks record takers to read the circumference of the tree from 1.3m off the ground it is rooted in.
-This measurement has been known to be vague, especially when it comes to warped or tipping trees, making the exact measurements made on each tree inaccurate at best.
-
-Using R and R Studio
+There is also measurement called dbh_trunk.
+This is the standard method forestry experts use to measure the diameter of a tree.
+Depending on where the measurement is taking placce, the breast height changes.
+For example, in Canada it is 1.3 meters, while the US uses 4.5 feet.
+This methodology is comprised of old conventions used to measure trees used throughout the world (@dbh).
+The result is a complicated mess where breast_height is a vague term that changes by location.
+Additionally, the convention breaks down when the tree is anything but vertically upright and straight.
+As a result, the DBH is more of an estimate rather than a sturdy measurement.
 
@@
 #| label: fig-5
-#| fig-cap: Distribution of Distance-at-Breast-Height diameter of trees in Toronto up to September 2024 (excluding trees with a recorded DBH of over 200)
+#| fig-cap: Distribution of Distance-at-Breast-Height diameter of all trees at or under 200 cm DBH in Toronto up to September 2024
 #| echo: false
@@
 ```
+@fig-5 is the result of trimming large dbh_trunk observations from @fig-4.
+This gives us a better idea of where the average dbh is, which is around 8-12 cm.
+However, this average would have been made without the large DBH values mentioned earlier, which would skew this average to a higher number.
+
+\newpage
 
@@
 
-## The supposed lack of trees in downtown Toronto
+While @fig-3 demonstrates that Acer platanoides isn't the most popoulous tree everywhere in Toronto, it also shows the potential dangers having non-native species have on places with little natural wildlife.
+Ward 13 has the fewest street trees in Toronto, and ther are still over 30 trees of a species known to be invasive and hamper the survival of native plants and trees.
+Additionally, since the downtown core is full of tall buildings which carry wind currents, the dispersal of Acer platanoides seeds seems inevitable, and will only exacerbate the troubling increasing in population of this invasive species, across Toronto and Southern Ontario.
+More care should be put into the planning of street trees, putting more empahsis on native trees in spaces with minial room for any trees to begin with.
+
+## The lack of trees in Toronto-Centre
 @fig-2 shows the distribution of trees across the 25 wards of Toronto.
@@
 
-## DBH diameters
+Knowing that Ward 13 has the lowest amount of street trees across all of Toronto's wards, @fig-3 was made to investigate the tree species count in this ward.
+Unlike the city-wide trend of Acer platanoides being the most popoulous, the most popoulous tree in this ward is Gleditsia triacanthos, otherwise known as the Honey Locust.
+The honey Locust.
+The HOney Locust is a tree that grows moderately fast and resiliant against dry and salty- climates.
+This makes it ideal for windbreaking (@blair1990gleditsia).
+Windbreaking, as mentioned earlier, helps create inviting spaces in urban enviroments, especially for pedestrians.
+Using a tree such as Honey Locut makes sense therefore, in a crowded downtore core with large skyscrapers and condos, which funnel air and wind around them.
+The species is native to North America, but it's reach just barely makes into the southernmost areas of Ontario, making it technically a non-native species in Toronto (@blair1990gleditsia).
 
-The spread of DBH diameters indicates that a small number of trees have extremely high DBH values.
+## Anomalies in DBH diameters and potential causes
+The spread of DBH diameters as seen in @fig-4 indicates that a small number of trees have extremely high DBH values.
 A DBH value of 800 for instance, means that the measured diameter of the tree is 8 meters!
@@
 There may have been a decimal missing from the entry, which would divide the offending dbh by 10.
-We can't be certain when the data was taken and how accurate it is however, so we leave it as is for future examination.
-The largest bin is located somewhere between 10 cm and 20 cm, indicating that the median diameter of Toronto's street trees is around that value as well.
+@fig-5 was made excluding any dbh_trunk values over 200 cm, to get a better look at the majority of all trees in Toronto.
+It shows a much clearer distribution of the vast majority of all dbh_trunk values in the data.
+The largest bin is located somewhere between 8 cm and 12 cm, indicating that the averaege diameter of Toronto's street trees is around that value as well.
 This suggests that most street trees are actually very young, and have not had much time to develop their trunks.
 This can be explained by initiatives to plant saplings along new or existing roads, potentially replacing older trees that had been cut down to make way for urban development.
+Larger trees are more rare either because they have been removed for urban developemnt, mentioned before, or as a result of storms and natural disasters.
 
@@
 As a result the data may be mismatched, with some trees having only been checked months or even years prior.
-Another limitation mentioned earlier is the inclusion of some privately-owned trees.
-There isn't a clear indicator mentioned of who owns the tree in each ovbservation.
+This may be why @fig-4 is so skewed.
+There may have been errors in the way the data was inputted for certain trees, leading to misplaced decimals.
+A similar occurrence happened when observing the raw data, which can be seen in the Appendix.
+No additional information was provided by OpenData Toronto when viewing the dataset on their database, apart from who provided the data, Urban Forestry.
+Perhaps more detail on their part, including units of measurement for dbh_trunk, would have helped make this analysis much easier.Another limitation mentioned earlier is the inclusion of some privately-owned trees.
 
-Because of the scope of the data harvested, there may be some inacuraccies.
-For example, Figure 3 is skewed extremely far to the left due to some enourmous numbers.
-There isn't a clear way of double cheking whther a dbh of 815 was actually recorded and accurate, which makes any conclusions drawn from DBH potentially inaccurate.
+Next steps would be augmenting this data with other data from Toronto's roads, or other trees, to see if any comparisions or correlations can be drawn there.
+Does vehicle traffic impact the trees lining a certain road, and if so, how long would it take for the effects to be noticeable?
+Does the distribution of street trees tell us anything about income disparity, fire risk, or quality of life within Toronto?
+This report can be used as a launching pad for further investigations, using key takeways from this report as drivers for further questions.
+
+With newer report, pressure will be put upon Urban Forestry to better curtail the data they have collected thus far.
+It is imperative that their hard work be recognized and any feedback on the data they have collected be sent to them as well, ensuring better datasets for analysts, leading to more informative conlusions for all.
 

### 2024-09-27 scripts/05-replications.R

@@ -40,7 +40,20 @@ ward_count |>
   labs(x = "Ward",
        y = "Number of trees planted")
 
-# Figure 3: Distribution of Distance-at-Breast-Height diameter of all trees in Toronto up to September 2024
+# Figure 3: Top 10 tree species count Toronto's Ward 13 (Toronto-Centre)
+ward13_species_count <- analysis_data %>%
+  filter(ward == "13") %>%
+  count(botanical_name, sort = TRUE) %>%
+  head(10)
+
+ward13_species_count |> 
+  ggplot(aes(y = botanical_name, x = n)) +
+  geom_col() +
+  theme_minimal() +
+  labs(x = "Planted Trees",
+       y = NULL)
+
+# Figure 4: Distribution of Distance-at-Breast-Height diameter of all trees in Toronto up to September 2024
 analysis_data |> 
   ggplot(mapping = aes(x = dbh_trunk)) +
   geom_histogram() +
@@ -48,4 +61,14 @@ analysis_data |>
   labs(x = "DBH Diameter",
        y = "Tree Count")
 
+# Figure 5: Distribution of Distance-at-Breast-Height diameter of all trees at or under 200 cm DBH in Toronto up to September 2024
+filter200_data <- analysis_data %>%
+  filter(dbh_trunk <= 200)
+
+filter200_data |> 
+  ggplot(aes(x = dbh_trunk)) +
+  geom_histogram(binwidth = 10) +
+  theme_minimal() +
+  labs(x = "DBH Diameter (cm)",
+       y = "Tree Count")
 

### 2024-09-27 scripts/05-replications.R
Added references to LLM usuage, final check and submission

@@ -21,6 +21,7 @@ top_species_count <- analysis_data %>%
   count(botanical_name, sort = TRUE) %>%
   head(10)
 
+# Gather top 10 tree species by count (LLM usage was used to troubleshoot this function)
 top_species_count |> 
   ggplot(aes(y = botanical_name, x = n)) +
   geom_col() +
@@ -33,6 +34,7 @@ ward_count <- analysis_data %>%
   count(ward, sort = TRUE) %>%
   head(10)
 
+# Gather counts of trees per ward
 ward_count |> 
   ggplot(aes(x = ward, y = n)) +
   geom_col() +
